---
title: "Specdata"
output: html_notebook
---

The function named 'pollutantmean' calculates the mean of a pollutant (sulfate or nitrate) across a specified list of monitors.
This function takes three arguments: 'directory', 'pollutant', and 'id'.
Directory is a character vector of length 1 indicating the location of the csv file.
Pollutant is a character vector of length 1 indicating the name of the pollutant
for which we will calculate the mean; either 'sulfate' or 'nitrate'.
ID is an integer vector indicating the monitor ID numbers to be used.
```{r}
pollutantmean <- function(directory, pollutant, id=1:332) {
    # create a list of files
    files <- list.files(directory, full.names=TRUE)
    
    # now create an empty dataframe
    my_df <- data.frame()
    for(i in id) {
        # loop through the files, rbinding them together
        my_df <- rbind(my_df, read.csv(files[i]))
    }
    
    # take pollutant column from 'dat' dataframe, remove any 'NA's and calculate mean
    mean <- mean(my_df[, pollutant], na.rm=TRUE)
    return(mean)
}

pollutantmean("data/specdata", "sulfate", 1:10)
```
```{r}
pollutantmean("data/specdata", 'nitrate', 70:72)
```
```{r}
pollutantmean("data/specdata", 'nitrate', 23)
```

The function named 'complete' reads a directory full of files and reports the number of completely observed cases in each data file.
The function returns a dataframe where the first column is the name of the file and the second column is the number of complete cases.
```{r}
complete <- function(directory, id=1:332) {
    # create a list of files
    files <- list.files(directory, full.names=TRUE)
    my_df <- data.frame(id=character(), nobs=character(), stringsAsFactors = FALSE)
    for(i in id) {
        complete_cases <- nrow(na.omit(read.csv(files[i])))
        complete_data <- c(i, complete_cases)
        my_df <- rbind(my_df, complete_data)
    }
    colnames(my_df, do.NULL=FALSE)
    colnames(my_df) <- c("id", "nobs")
    return(my_df) # this will return the entire dataframe to the function caller
}

complete("data/specdata", 1)
```
```{r}
complete("data/specdata", c(2, 4, 8, 10, 12))
```

```{r}
complete("data/specdata", 30:25)
```
```{r}
complete("data/specdata", 3)
```


The function named 'corr' takes a directory of data files and a threshold for complete cases and calculates the correlation between sulfate and nitrate for monitor locations where the number of completely observed cases (on all variables) is greater than the threshold. The function returns a vector of correlations for the monitors that meet the threshold requirement. If no monitors meet the threshold requirement, then the function returns a numeric vector of length 0.

```{r}
corr <- function(directory, threshold=0) {
    # create a list of files
    files <- list.files(directory, full.names=TRUE)
    # create a numeric vector that is empty
    vect <- vector(mode="numeric", length=0)
    comp_total <- list()
    # first find out if each file has at least the threshold number of complete observations
    for(i in seq_along(files)) {
        comp <- complete(directory, i)
        comp_total <- c(comp_total, list(comp$nobs))
        if(comp$nobs >= threshold) {
            my_file <- read.csv(files[i])
            complete_files <- na.omit(my_file)
            correlation <- cor(complete_files$sulfate, complete_files$nitrate)
            # append each correlation to the vector
            vect <- c(vect, correlation)
        }
    }
    return(vect)
}

cr <- corr("data/specdata", 150)
head(cr)
```

```{r}
summary(cr)
```
```{r}
cr <- corr("data/specdata", 400)
head(cr)
```

```{r}
summary(cr)
```

```{r}
cr <- corr("data/specdata", 5000)
head(cr)
```
```{r}
summary(cr)
```
```{r}
length(cr)
```
```{r}
cr <- corr("data/specdata")
summary(cr)
```

```{r}
length(cr)
```


```{r}
pollutantmean("data/specdata", "sulfate", 34)
```

```{r}
pollutantmean("data/specdata", "sulfate")
```
```{r}
cc <- complete("data/specdata", c(6, 10, 20, 34, 100, 200, 310))
print(cc$nobs)
```
```{r}
set.seed(42)
cc <- complete("data/specdata", 332:1)
use <- sample(332, 10)
print(cc[use, "nobs"])
```

```{r}
cr <- corr("data/specdata")
cr <- sort(cr)
set.seed(868)
out <- round(cr[sample(length(cr), 5)], 4)
print(out)
```

```{r}
cr <- corr("data/specdata", 129)                
cr <- sort(cr)                
n <- length(cr)                
set.seed(197)                
out <- c(n, round(cr[sample(n, 5)], 4))
print(out)
```
```{r}
cr <- corr("data/specdata", 2000)                
n <- length(cr)                
cr <- corr("data/specdata", 1000)                
cr <- sort(cr)
print(c(n, round(cr, 4)))
```

